---
title: "Term Structure Models"
author: "Caleb Vinson"
date: " "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(ggplot2)

for (.f in sort(list.files("R", pattern = "\\.R$", full.names = TRUE))) source(.f)
```

The theory behind each model below -- bootstrapping, the risk-neutral discount
formula, the affine SDE family, and where each named model sits in it -- is
written up in [docs/THEORY.md](docs/THEORY.md). This file is the narrative and
the code; the full numeric validation this narrative draws on is regenerated
by `Rscript run.R validate` into [docs/VALIDATION.md](docs/VALIDATION.md), and
the "What was wrong" section in the README documents what changed from the
original version of this notebook (kept at
[legacy/Term_Structure_Discussions_original.Rmd](legacy/Term_Structure_Discussions_original.Rmd)).

```{r model-comparison, echo=FALSE}
models <- data.frame(
  Model = c("Vasicek", "CIR", "Hull-White", "Black-Karasinski", "Nelson-Siegel"),
  Uniqueness = c(
    "Mean-reverting with constant volatility.",
    "Non-negativity via square-root volatility.",
    "Time-dependent mean reversion, fit to today's curve.",
    "Logarithmic modeling of the short rate.",
    "Empirical yield curve representation."
  ),
  Pro = c(
    "Analytical tractability.",
    "Ensures positive rates.",
    "Fits initial yield curve exactly.",
    "Ensures positive rates.",
    "Intuitive interpretation of parameters."
  ),
  Con = c(
    "Allows negative rates.",
    "Difficult to calibrate.",
    "Increased calibration complexity.",
    "No closed-form bond price.",
    "Lacks arbitrage-free properties."
  )
)

kable(models, caption = "Comparison of interest rate models") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

## Short rate models

Each model is simulated over many paths rather than one, which is the point:
a single path at a single seed says nothing about the distribution a model
implies, only about the one draw that happened to come up. `R/short_rate.R`
has the full documentation for how non-negativity is enforced for CIR and
Black-Karasinski without the clipping bias the original notebook had.

```{r vasicek, fig.cap="Vasicek: 30 paths, r0=0.03, a=0.5, b=0.04, sigma=0.015"}
v <- simulate_vasicek(r0 = 0.03, a = 0.5, b = 0.04, sigma = 0.015, Tt = 1, dt = 1 / 252,
                       n_paths = 30, seed = 123)
df <- data.frame(t = attr(v, "times"), v)
df_long <- reshape(df, varying = names(df)[-1], v.names = "rate", direction = "long",
                    timevar = "path")
ggplot(df_long, aes(t, rate, group = path)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  labs(title = "Vasicek short rate, 30 paths", x = "time (years)", y = "short rate") +
  theme_minimal()
```

```{r cir, fig.cap="CIR: 30 paths via the exact noncentral chi-square scheme"}
cir <- simulate_cir(r0 = 0.03, a = 0.8, b = 0.05, sigma = 0.12, Tt = 1, dt = 1 / 252,
                     n_paths = 30, seed = 123, method = "exact")
stopifnot(all(cir >= 0))
df <- data.frame(t = attr(cir, "times"), cir)
df_long <- reshape(df, varying = names(df)[-1], v.names = "rate", direction = "long",
                    timevar = "path")
ggplot(df_long, aes(t, rate, group = path)) +
  geom_line(alpha = 0.3, color = "darkgreen") +
  labs(title = sprintf("CIR short rate, 30 paths (min observed: %.4f)", min(cir)),
       x = "time (years)", y = "short rate") +
  theme_minimal()
```

```{r black-karasinski, fig.cap="Black-Karasinski: 30 paths, log-rate Euler"}
bk <- simulate_bk(r0 = 0.03, a = 0.5, theta = 0.04, sigma = 0.25, Tt = 1, dt = 1 / 252,
                   n_paths = 30, seed = 123)
stopifnot(all(bk > 0))
df <- data.frame(t = attr(bk, "times"), bk)
df_long <- reshape(df, varying = names(df)[-1], v.names = "rate", direction = "long",
                    timevar = "path")
ggplot(df_long, aes(t, rate, group = path)) +
  geom_line(alpha = 0.3, color = "brown") +
  labs(title = "Black-Karasinski short rate, 30 paths", x = "time (years)", y = "short rate") +
  theme_minimal()
```

## Hull-White, fit to today's curve

Unlike the three models above, Hull-White's mean-reversion level theta(t) is
not a constant: it is derived from the cached Treasury curve
(`data/treasury_curve.csv`, loaded through `R/data.R`) so that the model
reproduces that curve exactly at t=0 by construction. See
`simulate_hull_white()` and `hull_white_bond_price()` for the formulas.

```{r hull-white}
fred <- load_treasury_curve()
dc <- discount_curve_from_zero_rates(fred$maturity_years, fred$yield_pct / 100)
r0 <- forward_rate_from_discount(dc)(0)

hw <- simulate_hull_white(r0, a = 0.5, sigma = 0.01, Tt = 1, dt = 1 / 252, dc,
                           n_paths = 30, seed = 123)
df <- data.frame(t = attr(hw, "times"), hw)
df_long <- reshape(df, varying = names(df)[-1], v.names = "rate", direction = "long",
                    timevar = "path")
ggplot(df_long, aes(t, rate, group = path)) +
  geom_line(alpha = 0.3, color = "purple") +
  labs(title = "Hull-White short rate, 30 paths", x = "time (years)", y = "short rate") +
  theme_minimal()
```

```{r hull-white-fit, echo=FALSE}
maturities_check <- c(1, 2, 5, 10, 20)
check <- data.frame(
  maturity = maturities_check,
  curve = sapply(maturities_check, dc),
  hull_white = sapply(maturities_check, function(Tt) {
    hull_white_bond_price(r0, a = 0.5, sigma = 0.01, t = 0, Tt = Tt, discount_curve = dc)
  })
)
kable(check, digits = 6, caption = "Hull-White reproduces the input curve exactly at t=0") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Nelson-Siegel

Fit to the same cached Treasury curve, rather than to six hardcoded points.

```{r nelson-siegel}
fit <- fit_ns(fred$maturity_years, fred$yield_pct / 100)
tau <- seq(0.01, 30, length.out = 300)
fitted_curve <- nelson_siegel(tau, fit$params$beta0, fit$params$beta1, fit$params$beta2,
                               fit$params$lambda)

ggplot() +
  geom_line(data = data.frame(tau = tau, yield = fitted_curve * 100),
            aes(tau, yield), color = "steelblue", linewidth = 1) +
  geom_point(data = data.frame(maturity = fred$maturity_years, yield = fred$yield_pct),
             aes(maturity, yield), color = "firebrick", size = 2) +
  labs(title = sprintf("Nelson-Siegel fit, %s (RMSE %.1f bp)", fred$date[1], fit$rmse * 1e4),
       x = "maturity (years)", y = "yield (%)") +
  theme_minimal()
```

Full numeric validation -- Monte Carlo vs closed-form bond prices, moment
checks, and the Nelson-Siegel parameter recovery on synthetic data -- is in
[docs/VALIDATION.md](docs/VALIDATION.md), regenerated by `Rscript run.R
validate`.
